\paperTitle

### 2014-05-12 doc/icml2014/abstract.tex
Various minor corrections to camera-ready
gotten up to lemma 2

Recent work on the method of moments provide consistent estimates
  for latent-variable models, avoiding local optima issues,
  but are applicable only to certain types of graphical models.
  On the other hand, pure likelihood objectives are difficult to optimize.
In this work, we show that using the method of moments in conjunction
  with composite marginal likelihood
  yields consistent parameter estimates for a much broader
  class of discrete directed and undirected graphical models,
  including loopy graphs with high treewidth.
%In particular, our algorithm applies to any graphical model wherein each
  %hidden variable has at least three conditionally independent observed
  %variables.
  %that sidestep local optima issues for a general class of
  %directed and undirected graphical models.
%Parameter estimation in latent-variable models based on EM
  %is prone to local optima.
  %due to the non-convexity of the marginal log-likelihood. 
%In this paper, we extend recent advances in the method of moments 
  %to provide consistent parameter estimates
  %that sidestep local optima issues for a general class of
  %directed and undirected graphical models.
%In particular, our algorithm applies to any graphical model wherein each
  %hidden variable has at least three conditionally independent observed
  %variables.
%This family includes loopy graphs with high treewidth.
Specifically, we use tensor factorization to reveal partial information
about the hidden variables.  This renders the otherwise non-convex
likelihood objective convex, and EM can then be used to attain the global optimum.
%We use the method of moments to tie each hidden to some observed
  %variables, and show that the rest of the problem can be cast as convex
  %optimization of the likelihood.
%Our approach gracefully extends to models outside our class by
  %incorporating the partial information via posterior regulraization.
\end{abstract}

>>>>

Recent work on the method of moments enable
  consistent parameter estimation, but
  only for certain types of latent-variable models. %, avoiding local optima issues,
  %but are applicable only to certain types of graphical models.
On the other hand, pure likelihood objectives, though more universally applicable,
  are difficult to optimize.
In this work, we show that using the method of moments in conjunction
  with composite marginal likelihood
  yields consistent parameter estimates for a much broader
  class of discrete directed and undirected graphical models,
  including loopy graphs with high treewidth.
%In particular, our algorithm applies to any graphical model wherein each
  %hidden variable has at least three conditionally independent observed
  %variables.
  %that sidestep local optima issues for a general class of
  %directed and undirected graphical models.
%Parameter estimation in latent-variable models based on EM
  %is prone to local optima.
  %due to the non-convexity of the marginal log-likelihood. 
%In this paper, we extend recent advances in the method of moments 
  %to provide consistent parameter estimates
  %that sidestep local optima issues for a general class of
  %directed and undirected graphical models.
%In particular, our algorithm applies to any graphical model wherein each
  %hidden variable has at least three conditionally independent observed
  %variables.
%This family includes loopy graphs with high treewidth.
Specifically, we first use tensor factorization to reveal information
about the hidden variables.  This renders the otherwise non-convex
likelihood objective convex, and EM can then be used to obtain the global optimum.
%We use the method of moments to tie each hidden to some observed
  %variables, and show that the rest of the problem can be cast as convex
  %optimization of the likelihood.
%Our approach gracefully extends to models outside our class by
  %incorporating the partial information via posterior regulraization.
\end{abstract}

### 2014-05-12 doc/icml2014/background.tex
gotten up to lemma 2

Let $\sG$ be a discrete graphical model with observed variables $X$ and
  hidden variables $H$.
Let $\Pa(a) \subset X \union H$ be the parents of a variable $a$ in a directed
  model, and 
let $\sN(a) \subset X \union H$ be the neighbors of $a$ in an
  undirected model.
We will use $x_1, x_2, \cdots, x_L \in X$ to denote observed variables and
  $h_1, h_2, \cdots, h_M \in H$ to denote hidden variables.
For an integer $n$, let $[n] = \{ 1, \dots, n \}$.
For ease of exposition, we assume that $x_v \in [d]$ for
all $v$, and that $h_i \in [k]$ for all $i$.
Let $\sX$ be the joint domain of the observed variables, i.e. $\sX \eqdef [d]^L$ and
let $\sH$ be the joint domain of the hidden variables, i.e. $\sH \eqdef [k]^M$.

% PL: don't know what this means
%We assume that the model $\sG$ is parametrized at every clique $\sC$.
If $\sG$ is directed, the model parameters correspond to
the conditional probability tables for each clique $(\Pa(a), a) \in \sG$.
If $\sG$ is undirected, we assume a log-linear parametrization of each clique $\sC \in \sG$.

@@

% Maximum likelihood
The standard estimation procedure is maximum (marginal) likelihood,
  \begin{align*}
    L_\text{unsup}(\theta) &\eqdef \sum_{\vx \in \sD} \log p_\theta(\vx)
        =  \sum_{\vx \in \sD} \log \sum_{\vh \in \sH} p_\theta(\vx,\vh).
  \end{align*}
  Maximum likelihood is statistically efficient,
  but in general computationally intractable
  because marginalizing over hidden variables $\vh$ yields a non-convex objective.
In practice, one uses local optimization procedures (e.g., EM
  or L-BFGS) on the marginal likelihood, which can get stuck in local
  optima.
We will later return to likelihoods, but let us first
describe a method of moments approach for parameter estimation.
The starting point is moments of the marginal distribution $\BP(\vx)$,
and a bit of notation is necessary.

@@

We use the notation $[\cdot]$ to indicate indexing; for example $M[i]$
  is the $i$-th row of a matrix $M$ and $M[i,j]$ is the $i,j$-th element
  of $M$.
For a tensor $T \in \Re^{d \times \ldots \times d}$ and a vector $\bi
  = (i_1, \cdots, i_\ell)$, define the projection $T[\bi] = T[i_1 \ldots
  i_\ell]$.
Similarly, for a set $\sC$ of size $m$ and vector $\bx \in \Re^d$, we
  will use the notation $\bx_\sC$ to get a vector in $\Re^m$ which selects
  the appropriate elements of $\bx$.

We use $\otimes$ to denote the tensor product: if $u \in \Re^d,
  v \in \Re^k$, then $u \otimes v \in \Re^{d \times k}$.
For an $\ell$-th order tensor $T \in \Re^{d \times \ldots \times
  d}$ and vectors $\bx_1, \cdots, \bx_\ell \in \Re^{d}$, define 
  the application:
\begin{align*}
  T(\bx_1, \cdots, \bx_\ell) 
      %&= \sum_{i_1 = 1}^{d} \cdots \sum_{i_\ell = 1}^{d} 
&= \sum_{\bi = \ones}^{\vec d}
            T[i_1, i_2, \ldots, i_\ell] \bx_1[i_1] \cdots \bx_\ell[i_\ell].
\end{align*}
Analogously, for a matrices $M_1 \in \Re^{d \times k}, \cdots,
  M_\ell \in \Re^{d \times k}$, we define
\begin{align*}
  T(M_1, \cdots, M_\ell)[\vec j]
      %&= \sum_{i_1 = 1}^{d} \cdots \sum_{i_\ell = 1}^{d} 
  &= \sum_{\bi}
            T[\vec i] {M_1}[i_1,j_1] \cdots M_\ell[i_\ell, j_\ell],
\end{align*}
a tensor in $\Re^{k \times \ldots \times k}$.
%where $\vec i = (i_1, \cdots, i_\ell)$ and $\vec j = (i_1, \cdots, j_\ell)$

We consider up to third-order moments:
\begin{align*}
  \mO_i &\eqdef \Pr(x_i), &
  \mO_{ij} &\eqdef \Pr(x_i, x_j), &
  \mO_{ijk} &\eqdef \Pr(x_i, x_j, x_k).
\end{align*}
We will view $\mO_i, \mO_{ij}, \mO_{ijk}$ as tensors of
  orders $1, 2, 3$ in $\Re^d, \Re^{d\times d}, \Re^{d \times d \times d}$.
Next, we define the \emph{hidden marginals}:
\begin{align*}
  \mH_i &\eqdef \Pr(h_i), &
  \mH_{ij} &\eqdef \Pr(h_i, h_j), &
  \mH_{ijk} &\eqdef \Pr(h_i, h_j, h_k).
\end{align*}
These are tensors of
  orders $1, 2, 3$ in $\Re^k, \Re^{k\times k}, \Re^{k \times k \times k}$.

@@

\citet{anandkumar12moments} presented a consistent estimator
  using the method of moments to learn the parameters of a three-view
  mixture model (\figureref{three-view}), an important case of directed
  graphical models. %, that sidesteps the local optima problem altogether. 
In that model, we have three observed variables $x_1, x_2, x_3$
  that are independent conditioned on a hidden variable $h_1$.  
The parameters of this model are the mixture probabilities $\pi\oft 1 \eqdef
  \Pr(h_1) \in \Re^k$ and the conditional moments $\mOpp{i}{1} \eqdef
  \Pr(x_i \given h_1) \in \Re^{d \times k}$ for $i \in \{1,2,3\}$.

@@

\begin{assumption}
\label{asm:full-rank}
The conditional moments $\mOpp{1}{1}$, $\mOpp{2}{1}$, $\mOpp{3}{1}$ have full column rank
  $k$, and that $\pi\oft 1 \succ 0$.
\end{assumption}

In brief, the algorithm first computes a whitening matrix $W \in
  \Re^{d \times k}$ such that $W^\top \mO_{12} W = I_{k \times k}$,
  and uses it to transform $\mO_{123}$ into a symmetric orthogonal tensor.  
Then a robust tensor power method is applied to extract the eigenvectors
of the whitened $\mO_{123}$; unwhitening yields the columns of $\mOpp{3}{1}$.
  $\mOpp{1}{1}$ and $\mOpp{2}{1}$ can then be recovered with a few matrix operations.

>>>>

Let $\sG$ be a discrete graphical model with observed variables $X$ and
  hidden variables $H$.
For a directed model,
  let $\Pa(a) \subset X \union H$ be the parents of a variable $a$;
  for an undirected model,
  let $\sN(a) \subset X \union H$ be the neighbors of $a$.
  Abusing notation slightly, we use $\sG$ to denote the set of cliques.
  %(for directed models, these are $\sG = \{ \Pa(a) \union \{a\} : a \in X \union H \}$).
We use $x_1, x_2, \dots, x_L \in X$ to denote observed variables and
  $h_1, h_2, \dots, h_M \in H$ to denote hidden variables.
For an integer $n$, let $[n] = \{ 1, \dots, n \}$.
For ease of exposition, we assume that $x_v \in [d]$ for
all $v \in [L]$, and that $h_i \in [k]$ for all $i \in [M]$.
Let $\sX$ be the joint domain of the observed variables, i.e. $\sX \eqdef [d]^L$, and
let $\sH$ be the joint domain of the hidden variables, i.e. $\sH \eqdef [k]^M$.

% PL: don't know what this means
%We assume that the model $\sG$ is parametrized at every clique $\sC$.
For directed models,
$p_\theta(\bx,\bh) = \prod_{a \in X \cup H} p_\theta(a \mid \Pa(a))$,
where the parameters $\theta$ are the conditional probability tables.
For undirected models,
$p_\theta(\bx,\bh) \propto \prod_{\sC \in \sG} \exp(\theta^\top\phi_\sC(\bx_\sC, \bh_\sC))$,
where $\theta$ are the canonical parameters of an exponential family.

% PL: log-linear parametrization - can't say this for each clique: this is a global notion.
%If $\sG$ is undirected, we assume a log-linear parametrization of each clique $\sC \in \sG$.

@@

% Maximum likelihood
The standard estimation procedure is maximum (marginal) likelihood,
  \begin{align*}
    L_\text{unsup}(\theta) &\eqdef \sum_{\vx \in \sD} \log p_\theta(\vx)
        =  \sum_{\vx \in \sD} \log \sum_{\vh \in \sH} p_\theta(\vx,\vh).
  \end{align*}
  Maximum likelihood is statistically efficient,
  but in general computationally intractable
  because marginalizing over hidden variables $\vh$ yields a non-convex objective.
In practice, one uses local optimization procedures (e.g., EM
  or L-BFGS) on the marginal likelihood, which can get stuck in local
  optima.
We will later return to likelihoods, but let us first
describe a method of moments approach for parameter estimation.
First, a bit of notation is necessary.
%The starting point is moments of the marginal distribution $\BP(\vx)$,
%and a bit of notation is necessary.

@@

We use the notation $[\cdot]$ to indicate indexing; for example $M[i]$
  is the $i$-th row of a matrix $M$ and $M[i,j]$ is the $i,j$-th element
  of $M$.
For a tensor $T \in \Re^{d \times \cdots \times d}$ and a vector $\bi
  = (i_1, \ldots, i_\ell)$, define the projection $T[\bi] = T[i_1, \ldots, i_\ell]$.
Similarly, for a set $\sC$ of size $m$ and vector $v \in \Re^d$, we
  will use the notation $v_\sC$ to get a vector in $\Re^m$ which selects
  the appropriate elements of $v$.

We use $\otimes$ to denote the tensor product: if $u \in \Re^d,
  v \in \Re^k$, then $u \otimes v \in \Re^{d \times k}$.
For an $\ell$-th order tensor $T \in \Re^{d \times \ldots \times
  d}$ and vectors $v_1, \cdots, v_\ell \in \Re^{d}$, define 
  the application:
\begin{align*}
  T(v_1, \cdots, v_\ell) 
      %&= \sum_{i_1 = 1}^{d} \cdots \sum_{i_\ell = 1}^{d} 
% PL: this isn't great notation; reader can figure it out
%&= \sum_{\bi = \ones}^{\vec d}
  &= \sum_{\bi}
            T[\bi] v_1[i_1] \cdots v_\ell[i_\ell].
\end{align*}
Analogously, for a matrices $M_1 \in \Re^{d \times k}, \cdots,
  M_\ell \in \Re^{d \times k}$, we define
\begin{align*}
  T(M_1, \cdots, M_\ell)[\vec j]
      %&= \sum_{i_1 = 1}^{d} \cdots \sum_{i_\ell = 1}^{d} 
  &= \sum_{\bi}
            T[\vec i] {M_1}[i_1,j_1] \cdots M_\ell[i_\ell, j_\ell].
\end{align*}
%a tensor in $\Re^{k \times \ldots \times k}$. % PL: not true, since [j] is applied
%where $\vec i = (i_1, \cdots, i_\ell)$ and $\vec j = (i_1, \cdots, j_\ell)$

%with respect to the true data distribution $p_{\theta^*}(\bx,\bh)$;
We will use $\Pr(\cdot)$ to denote various moment tensors
constructed from the true data distribution $p_{\theta^*}(\bx,\bh)$:
%We consider up to third-order moments:
\begin{align*}
  \mO_i &\eqdef \Pr(x_i), &
  \mO_{ij} &\eqdef \Pr(x_i, x_j), &
  \mO_{ijk} &\eqdef \Pr(x_i, x_j, x_k).
\end{align*}
Here, $\mO_i, \mO_{ij}, \mO_{ijk}$ are tensors of
  orders $1, 2, 3$ in $\Re^d, \Re^{d\times d}, \Re^{d \times d \times d}$.
Next, we define the \emph{hidden marginals}:
\begin{align*}
  \mH_i &\eqdef \Pr(h_i), &
  \mH_{ij} &\eqdef \Pr(h_i, h_j), &
  \mH_{ijk} &\eqdef \Pr(h_i, h_j, h_k).
\end{align*}
These are tensors of
  orders $1, 2, 3$ in $\Re^k, \Re^{k\times k}, \Re^{k \times k \times k}$.

@@

\citet{anandkumar12moments} presented a consistent estimator
  using the method of moments to learn the parameters of a three-view
  mixture model (\figureref{three-view}), an important case of directed
  graphical models. %, that sidesteps the local optima problem altogether. 
In that model, we have three observed variables $x_1, x_2, x_3$
  that are independent conditioned on a hidden variable $h_1$.  
The parameters of this model are the prior probabilities $\pi\oft 1 \eqdef
  \Pr(h_1) \in \Re^k$ and the conditional moments $\mOpp{i}{1} \eqdef
  \Pr(x_i \given h_1) \in \Re^{d \times k}$ for $i \in \{1,2,3\}$.

@@

\begin{assumption}
\label{asm:full-rank}
The conditional moments $\mOpp{1}{1}$, $\mOpp{2}{1}$, $\mOpp{3}{1}$ have full column rank
  $k$, and $\pi\oft 1 \succ 0$.
\end{assumption}

In brief, the algorithm first computes a whitening matrix $W \in
  \Re^{d \times k}$ such that $W^\top \mO_{12} W = I_{k \times k}$,
  and uses $W$ to transform $\mO_{123}$ into a symmetric orthogonal tensor.  
Then a robust tensor power method is applied to extract the eigenvectors
of the whitened $\mO_{123}$; unwhitening yields the columns of $\mOpp{3}{1}$.
The other parameters can be recovered with a few additional matrix operations.
%$\mOpp{1}{1}$ and $\mOpp{2}{1}$ can then be recovered with a few matrix operations.
